---
title: How periodicDNA works
author: Jacques Serizay
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{How periodicDNA works}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
suppressPackageStartupMessages({
    library(GenomicRanges)
    library(magrittr)
    library(ggplot2)
    library(periodicDNA)
})
```

## Internal steps of periodicDNA

To estimate the periodicity strength of a given k-mer in one or several
sequences, periodicDNA does the following steps: 

1. It first finds all the possible k-mer occurrences over the sequence(s);
2. It then find all the possible pairwise distances and generates a 
   histogram of these pairwise distances; 
3. It then normalizes this histogram for distance-decay and smooths 
   the distribution to remove noise;
4. Finally, it performs a Fast Fourier Transform of the normalised histogram.

```{r, eval = TRUE, echo=FALSE, out.width='100%'}
knitr::include_graphics(
   "https://raw.githubusercontent.com/js2264/periodicDNA/master/man/figures/periodicityDNA_principle.png"
)
```
