---
title: periodicDNA
author: Jacques Serizay
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Introduction to periodicDNA}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", warning=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
suppressPackageStartupMessages({
    library(GenomicRanges)
    library(magrittr)
    library(ggplot2)
    library(periodicDNA)
})
```

## Introduction 

Many oligonucleotides are periodically occurring in genomes 
across eukaryotes, and some are impacting the physical properties of 
DNA. For example, DNA bendability is modulated by 10-bp periodic 
occurrences of WW (W = A/T) dinucleotides. periodicDNA is an R 
package offering a set of functions to identify k-mer periodicity 
in sequences of interest such as regulatory elements. 
It also provides a method to generate continuous tracks of k-mer 
periodicity over genomic loci. 

## Installation

periodicDNA can be installed from GitHub as follows:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("js2264/periodicDNA")
library(periodicDNA)
```

## Loading periodicDNA and other useful packages

```{r}
library(periodicDNA)
library(magrittr)
library(ggplot2)
```

## How periodicDNA works

When running `getPeriodicity()` function for a given k-mer (typically a
dinucleotide) over a set of GRanges, R does the following steps: 

1. It first finds all the possible motif occurrences over the GRanges;  
2. It then find all the possible pairwise distances and generates a 
   normalised histogram of the pairwise distances;  
3. It performs a Fast Fourier Transform of the normalised histogram; 
4. Finally, it outputs the power spectrum density of the k-mer in the 
   provided GRanges.

## Quick k-mer periodicity

A genome sequence and a set of genomic loci must be defined. 
Let's focus on the C. elegans genome for now, and more specifically 
around its ubiquitous TSSs. 

```{r, fig.wide = TRUE}
data(ce_proms)
ubiq_TSSs <- ce_proms[ce_proms$which.tissues == 'Ubiq.'] %>%
    deconvolveBidirectionalPromoters() %>% 
    alignToTSS(50, 300)
seqs <- getSeq(
    BSgenome.Celegans.UCSC.ce11::BSgenome.Celegans.UCSC.ce11
)
ubiq_seqs <- seqs[ubiq_TSSs]
ubiq_TT <- getPeriodicity(
    ubiq_seqs,
    motif = 'TT', 
    cores = 1
)
plotPeriodicityResults(ubiq_TT)
``` 

`getPeriodicity` can also works with GRanges objects, as long as a `genome`
is supplied. 'ce11', 'sacCer3', 'mm10', 'dm6', 'hg38' and 'danRer10' are 
currently supported. Otherwise, a BSgenome can be used directly. 

```{r, fig.wide = TRUE}
ubiq_TT <- getPeriodicity(
    ce_proms, 
    genome = 'ce11', 
    motif = 'TT', 
    cores = 1
)
``` 

<!--  
## Fold Power Increase metric

The FPI (fold power increase) metric was originally described in 
Pich et al., Cell 2018. It allows to accurately estimate the periodicity 
strength of a k-mer of interest by modelling the background periodicity.

```{r, fig.small = TRUE}
fpi <- getFPI(
    ubiq_seqs,
    motif = 'TT', 
    parallel_shuffling = 1, 
    n_shuffle = 10
)
plotFPI(fpi)
``` 

**Note:** To calculate the FPI of a k-mer in a set of sequences, the sequences
of interest are first randomly shuffled (`n_shuffle` times). This process is
stochastic and for this reason, the FPI score might vary each time the 
`getFPI()` function is ran. To ensure accurate estimation of the FPI score, 
it is recommended to run the `getFPI()` function with a high `n_shuffle`. The 
process can be parallelised using the `parallel_shuffling` argument. 

-->

## Generating periodicity tracks with periodicDNA

Rather than computing the periodicity of a chosen k-mer in a given 
set of sequences, `generatePeriodicityTrack()` function computes 
a linear periodicity track covering loci of interest. 

```{r}
generatePeriodicityTrack(
    seqs,
    granges = ce_proms[1], 
    motif = 'TT',
    freq = 1/10,
    cores = 1
)
```

```{r, echo=FALSE, results="hide", warning=FALSE}
unlink('TT-periodicity_g-100^2_b-100^5.bw')
```

## Session info

```{r}
sessionInfo()
```